\subsection*{Acknowledgements}

This research was funded by the National Institute for Health Research (NIHR), and supported by the NIHR Applied Research Collaboration (ARC) South West Peninsula. The views expressed are those of the authors and not necessarily those of the NHS, the NIHR or the Department of Health and Social Care.

We would like to thank the SAMueL project team (Lauren Asare, Leon Farmer, Julia Frost,  Iain Lang, Kristin Liabo, Peter McMeekin, Keira Pratt Boyden, Cathy Pope, Ken Stein, Penny Thompson) for their input into this work.

We would also like to thank our extended Patient and Carer Involvement team (David Burgess, Simon Douglas, Ian Hancock, Nicola Hancock, John Williams), and our expert advisory group (Ajay Bhalla, Gary Ford, Tony Rudd, and Martin Utley).

\subsection{Data access and ethics}

The NHS Health Research Authority decision tool\footnote{http://www.hra-decisiontools.org.uk/research/} was used to confirm that ethical approval was not required to access the data. No identifiable patient or hospital information were provided in the data, and anonymised hospital names were provided. Governance of the data and access to SSNAP was authorised by the Healthcare Quality Improvement Partnership\footnote{https://www.hqip.org.uk/} (HQIP, reference HQIP303). 

\subsection*{Funding}

This report is independent research funded by the National Institute for Health Research Applied Research Collaboration South West Peninsula and by the National Institute for Health Research Health and Social Care Delivery Research (HSDR) Programme [NIHR134326]. The views expressed in this publication are those of the authors and not necessarily those of the National Institute for Health Research or the Department of Health and Social Care.

\subsection*{Conflict of interest statement}

All authors declare no conflicts of interest.

\subsection*{Links to source code, online Jupyter book, and LaTeX paper}
\begin{enumerate}
    \item GitHub repository: \url{https://github.com/samuel-book/samuel_shap_paper_1} (Release 3.0.0; DOI: 10.5281/zenodo.7856495)
    \item Jupyter book: \url{https://samuel-book.github.io/samuel_shap_paper_1/}
    \item \LaTeX paper: \url{https://github.com/samuel-book/overleaf_shap_paper_1_short}
\end{enumerate}
